### 2020-12-27 response/response-demography-4.tex

\noindent I have attached the revised version of my paper,
``Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective
Abortions: India's Experience Over Four Decades.''
Thank you for the comments and suggestions. 
Below I outline my response to each issue raised.

The author has undertaken much reorganization and has made very useful clarifications to
benefit the reader in the text and tables. Nevertheless, we still find three issues with
the writing style:

(1) Explanations of key concepts or theories need to be spelled out more clearly,
particularly for the non-economist. Much has been done in this regard, but further work is
still needed.

RESPONSE: With the reorganization of the Conceptual Framework section described below, I now 
have an early paragraph that explains income and substitution effects in detail and how they 
tie into fertility decisions with increasing wages for both males and females.
I then rely on this detailed explanation for my later discussions of female education and
labor force participation.
Furthermore, I have provided additional explanations for decisions on investment in children 
based on the expected return to investment in their education.
Finally, I have reorganized my discussion of increased demand for mothers with more education
even in the presence of declining female labor force participation to make it clearer for
non-economists.

@@

(2) The arguments do not flow from section to section and paragraph to paragraph. Each
section and argument need to be introduced so the reader understands why this information
is being presented in this place. Many times, the author jumps into new arguments without
an introduction or transition or explanation.

RESPONSE: As detailed below, I have completely reorganized the Conceptual Framework section
to create a better flow and make it easier for the reader to follow.
Furthermore, I have added introductory paragraphs to the sections that did not already 
have these or an equivalent transition ("Conceptual Framework", "Estimation Strategy", "Data", 
and "How Birth Spacing Changed").
I have also rewritten paragraphs so they have the main argument first, followed by details
on the argument, which should make it easier to follow the thread of the arguments.

@@

(3) There is no clear outline of the arguments presented in the Background section. Each
argument is presented separately rather than tying into the overall framework and
subsequent analyses.

RESPONSE: I have completely reorganized the Conceptual Framework section to address this and
and both Issues (1) and (2) and the examples below.
It now begins with an overview of the section, followed by three potential explanations that
link fertility and birth spacing decisions: economic conditions, investment in children, and 
son preference.
I then introduce sex composition of previous children, female education, and area of
residence as the primary explanatory variables that can both proxy for the explanations and
that are available in the data.
Finally, I tie each of the three primary explanatory variables to the explanations and
discuss their predicted effects based on the potential explanations.

@@

The following are some examples of (1) and (2); however, the author should check the entire
paper for similar issues and address (3) as well.

1. In general, each argument in the background section is not clear and fully described for
the reader.

a. P. 4 first sentence: The author states that female education is a crucial explanatory
variable and it matters for fertility, sex ratios, mortality, etc. in a list. It would be
useful for the reader to briefly spell out how female education fits into the entire
framework looking across all these outcomes. In addition, the second paragraph on
measurement of education seems like it belongs in the data section, unless there is some
conceptual reason why this information is included here.

RESPONSE: I now introduce female education as a primary explanatory variable after the 
potential explanations linking fertility and birth spacing.
This allows me to better show how female education fits into the explanations and the potential 
effects of education on birth spacing.
I have moved the definitions of education groups to the Data section.

@@

b. The 2 paragraphs on female labor force participation on pp. 5-6: The two findings that
support the argument that changes in birth spacing are not due to economic reasons are
merely listed, leaving the reader to make the connection as to why these facts support the
argument herself. The income effect dominating the substitution effect and greater negative
elasticity of women’s wages need to be explained to readers unfamiliar with these concepts
or how they fit together.

RESPONSE: As detailed above, I have reorganized the whole section and now introduce income
and substitution effects early and apply these to fertility decisions.
This change should make it easier to understand the new explanation for why we are unlikely
to see a shortening of birth intervals in India even with increasing female wage.
I have removed the discussion of elasticity (but kept the reference) since this is simply
another way of saying that the income effect dominates the substitution effect.

@@

c. P. 6 regarding sex selection: “The introduction of sex selection allows parents to avoid
giving birth to girls but increases the expected interval to the next birth.” The reader
needs an explanation here of why sex selection increases birth interval length. The author
mentions waiting time to conception and other components in the introduction (and
footnote). A clear explanation of this argument needs to be reiterated (or moved) here.

RESPONSE: I have moved the original footnote into the text in the Conceptual Framework and
expanded the discussion to make it easier to understand.
The Introduction now simply says that it takes 6 months or more after an abortion to reach 
the same point in the next pregnancy and refers to the detailed discussion in the
Conceptual Framework section.

@@

d. P. 6: “As women’s education increases, their productivity in the production of offspring
human capital also increases. With relatively more boys born because of increased access to
sexselective abortions and the increasing income potential for (male) offspring, demand for
bettereducated women can increase, even if they do not participate in the labor market.” I
am unclear as to why this argument is included. It needs explanation. Is this related to
length of birth intervals here? Or is this mentioned because women’s education and
increasing demand for better-education women will actually lower the sex ratio?

RESPONSE: Yes, that was a terrible explanation.
The argument is exactly the opposite: the increased demand for better-educated women
is associated with a more male-biased sex ratio.
With declining female labor force participation it is clear that educated women are not
desired because they bring income to the household.
However, if better-educated women produce better-educated sons this may increase the
demand for better-educated women given the strong son preference and the rapid increasing 
return to male education.
The result is that higher female education may be associated with lower fertility, longer 
spacing, and increased use of sex selection even if the standard economic explanation
based on wages does not hold.
This is important because the usual policy recommendation of increased education may
not change the use of sex selection in the absence of changes in the relative return
to male and female education in the labor market.
I have completely rewritten the explanation and refer to it when discussing better
job opportunities in the Conclusion.

@@

e. Pp. 6-7: I do not understand the importance of this argument and how it fits into the
authors’ story: “If more and “better” parental attention per child results in higher child
“quality,” we should expect longer birth intervals. However, the evidence on spacing’s
effect on child quality measures such as IQ and education is mixed for developed countries
and nonexisting for developing countries. The exception is health and mortality, where
longer spacing does lead to better outcomes, although this relationship weakens with
maternal education.”

RESPONSE: The idea was to show that longer spacing could be thought of as an investment
in children that may occur at the same time as lower fertility.
I have completely rewritten that section to make this clearer.

@@

f. P. 7: The final summary paragraph lists predictions based on the earlier arguments. Yet
it is hard to connect these predictions to the paragraphs above. Perhaps predictions could
be noted with the earlier arguments.

RESPONSE: With the new structure of the Conceptual Framework section, the final paragraph
with the summary of predictions now follows directly the predictions for the three primary
explanatory variables.
Hence, it should connect better to the individual predictions.

@@

2. Estimation strategy section.

In the first paragraph, it would be useful to clearly outline the analyses that will be
undertaken (such as first, I document changes in birth intervals over time and how
influenced by sex selection, second how birth intervals affected fertility, and third how
birth intervals affected mortality). As it stands, this section jumps into a discussion of
the hazard model and the reader does not know in which analyses it will be used.

RESPONSE: I have added an outline of the three parts of the empirical analysis at the 
beginning of the section and highlighted that the Estimation Strategy section focuses
on the empirical model for birth spacing.

@@

Some additional issue as examples where more clarification is needed:

3. Abstract. Please reorganize to start with a statement of the problem and main aims of
the study. As it stands, the abstract is a list of findings, with no context as to the
issues/questions the author investigates. Please be cognizant of using terms such as “women
most likely to use sexselective abortion.” The reader does not understand how this is
measured yet, and may get the false impression that sex-selection abortion is measured
directly.

RESPONSE: I now begin the abstract with the changing sex ratio after the introduction 
of prenatal sex-determination technologies.
I then state when main issue, which is that abortions increase birth spacing but we
know little about how birth spacing has changed or the effects of these changes.
After this, I highlight the main birth spacing results.
I next discuss how the rapid lengthening of birth spacing led the period
fertility measure to substantially overestimate the decline in cohort fertility.
Finally, I disucss the question of whether repeated abortions can counteract the
other positive effects of longer birth spacing on infant mortality.

Throughout the abstract I have avoided ascribing use of sex-selection, instead refering to
the introduction of prenatal sex-determination technologies and the presumed use 
of sex-selective abortions.

@@

4. P. 1: The second motivation is unclear. “the combined changes in birth spacing may
outpace what we have observed in other countries.” Outpaced means what? decreasing faster,
or increasing faster, or sex selection plays more of a role?

RESPONSE: That was meant to capture the combined effect from more sex selection and the other changes.
I have rewritten the paragraph to emphasize that we do not know how much of an effect the 
combined changes will have on birth spacing and to remove the imprecise "outpace".
The new sentence is "The combination of the apparent increasing 
use of sex selection and the other societal changes can significantly impact birth spacing, 
but we know little about how much."

@@

5. Throughout the paper, the author compares the total fertility rate to cohort fertility
rate. For the former, is the period fertility rate a more accurate definition?

RESPONSE: I would argue that the total fertility rate is our most commonly used period
fertility measures, especially in "public-facing" discussions of changes in fertility 
behavior (see also the discussion in \citet{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}).
However, the make this clearer I have clarified whether I am referring to period fertility
or to cohort fertility throughout the paper.

@@

6. P. 2: the main point of this sentence is unclear: “Counteracting effect is possible if
longer birth intervals arise from multiple abortions because the short duration between
pregnancies could increase mortality.”

The idea is that even though the spacing between births become longer with sex-selective 
abortions, the spacing between \emph{pregnancies} may still be very short as suggested by a 
prior reviewer. 
If short pregnancy spacing has negative effects on child outcomes, the short pregnancy spacing 
may dampen or completely eliminate any positive effects from longer spacing. 
I have expanded the discussion of the possible counteracting effects in the Introduction.

The new paragraph is:
\begin{quote}
Second, what is the relationship between infant mortality and the changes in birth spacing
and sex selection? 
In India, birth intervals have traditionally been shorter with fewer sons, contributing to 
girls' higher mortality risk
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}. 
Therefore, longer birth spacing may reduce mortality through, for example, diminished 
sibling competition \citep{Conde-Agudelo2012,Molitoris2019}. 
However, if the spacing between births lengthens because of sex-selective abortions, the 
spacing between pregnancies may still be very short. 
Short pregnancy spacing may lead to worse child outcomes because of maternal nutritional 
depletion and insufficient time to recover from the previous pregnancy. 
Hence, children born after long birth intervals where multiple abortions have punctuated 
those intervals may not see the same benefits as children born after a long interval that 
is not punctuated by abortions.
\end{quote}

@@

7. Sometimes the use of the term “birth interval” actually refers to “birth interval
length.” Please check the uses of these terms and correct accordingly.

RESPONSE: Thank you for pointing this out, that was not caught by the copy editor
or me.
I have corrected this 28 diferent places in the paper.

@@

8. P. 2: “The key variables are maternal education, the sex of previous children, and the
area of residence.” Does the author mean key independent or predictor variables?

RESPONSE: Yes, those are the key explanatory variables. 
I have added that to the Introduction.
If you think that "independent" or "predictor" variables would be better, I would
be happy to change that throughout.

@@

9. The author should clearly note that sex selection is not observed and tone down some
conclusions, such as “Sex selection, however, is behind the most substantial increases in
birth spacing. The best-educated women with two girls had the most biased sex ratio and the
most significant increase in birth intervals.” The author could say that sex selection
appears to be behind these findings; or evidence suggests that sex selection is behind
these findings, etc. In addition, clarify that those most likely to practice sex selection
(women with highest education and two girls) is based on your analysis, not settled in the
literature.

RESPONSE:  To address this, I have made the following changes:
\begin{itemize}
\item Rewritten the Abstract as detailed above
\item In the Introduction's summary of results I lead with the changes in birth spacing and 
sex ratios and then say that these are likely arising from sex-selective abortions.
\item Specify that the grouping into more and less sex selection when analyzing birth spacing 
is based on sex ratios and clarify that the use of sex selection is inferred rather than 
established when analyzing birth spacing.
\item For the Conclusion, tone down the conclusions to focus on lengthening of birth intervals and 
that these likely came from sex selection as indicated by the increasingly male-biased sex ratio.
\end{itemize}

@@

10. While it might be messy to show the confidence intervals in all the figures, is it
possible to note some significant differences in the text?

RESPONSE: I did try again to come up with a way to incorporate confidence intervals in the figures,
but the only way that would work would triple the number of figures.
Instead, I have expanded the discussion of results to include where there are statistically
significant differences:
\begin{itemize}
\item I now highlighted that tests for statistically significant differences across sex compositions 
are available in the online appendix.
\item In the section on how birth spacing changed, I have added discussions of statistical significance 
when discussing comparisons across sex compositions.
\item In the Conclusion, I have also added that when there are no sons, we now see statistically 
significant longer, rather than statistically significant shorter, birth intervals than if there 
is at least one son, rather than simply discussing the reversal.
\end{itemize}

### 2021-01-08 paper/sex_selection_spacing-ver5.tex
Edits throughout the paper; mostly minor

\noindent Following the arrival of prenatal sex-determination technologies in the mid-1980s, 
India has experienced an increasingly male-biased sex ratio at birth, presumably from sex-selective 
abortions. 
Abortions lengthen birth intervals, but we know little about how birth spacing has changed or the 
effects of these changes. 
This paper shows that, although the length of birth intervals increased overall, well-educated women 
with no sons had the most substantial lengthening---and the most male-biased sex ratios. 
Furthermore, most of these changes took place immediately after the introduction of prenatal 
sex-determination technologies. 
Consequently, some women without sons now have longer birth intervals than those with sons, reversing 
India's traditional spacing pattern. 
Less-educated women continue short birth spacing when they have no sons, with only limited evidence 
of male-biased sex ratios. 
Because of the rapid lengthening of birth intervals, period fertility rates substantially overestimated 
how fast cohort fertility fell. 
Moreover, predicted cohort fertility is still 10\%--20\% above the period fertility rate. 
If lengthening of birth intervals occurs from repeated abortions, the associated short pregnancy spacing 
may counteract any positive effects of longer birth spacing. 
There is, however, no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.

@@

An increased desire to invest in children may also lengthen birth spacing, provided it has 
positive effects on child outcomes. 
The clearest example of a positive effect of longer spacing is for health as very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

@@

Despite that higher potential wages from more education should shorten birth spacing and
lower fertility, the low and declining female labor force participation suggests little
economic incentive to space children closer together.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

@@

However, higher female education might still be associated with lower fertility and more 
sex selection, provided better-educated women are better at producing child human capital, 
as suggested by \citet{Behrman1999}. 
In a situation with increasing returns to male education, households may desire 
better-educated women because of their ability to produce better-educated sons, 
rather than because of their higher potential income. 
The higher return to male education provides an additional incentive to invest in sons' 
education---with a better-educated mother one such investment---and leads to lower desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.

@@

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated as their
household income increase the most, and because of their use of sex selection. 
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

@@

Second, even without sex selection, it is unlikely that characteristics, such as 
the sex composition of previous births, have the same effects throughout the entire 
birth interval. 
The proportional hazard model requires that the hazard for any individual is a 
fixed proportion of the hazard for any other individual. 
Nonconstant effects violate that assumption, and the results from a proportional hazard 
model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals because there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.

@@

Therefore, I use a nonproportional hazard specification that allows the shape of the 
hazard functions to vary across groups.
The use of a nonproportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

I divide the explanatory variables into two groups, nonproportional and proportional. 
The first group consists of characteristics shown in the prior literature to affect the 
spacing choice and the use of sex selection: Mother’s education, sex composition of 
previous children, and area of residence.
To minimize any potential bias from including proportional variables, I estimate a 
separate model for each birth interval, education group, and period combination, rather 
than including education as a variable.

@@

Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentile birth interval durations in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with 12 or more years of 
education, and the fourth spell for women with 12 or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2\% boys
\citep{ben-porath76b,jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors and tests for statistically
significant differences across sex compositions are available in the online appendix.

@@

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth interval durations, 
where the more the parity progression probabilities declined, the more the birth interval 
lengthened. 
For example, the probability of a fourth birth for urban women with 8--11 years 
of education and two sons and one girl has declined by almost 40 percentage points as the 
75th percentile birth interval length increased by 22 months.
Compare this with rural women of no education with a boy as their first child, for whom the 
probability of a third birth declined by fewer than six percentage points while the birth 
interval increased only slightly over two months.

@@

Even more striking is that most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These birth spacing changes may even be underestimated because this particular group
appears to have had access to sex selection even before it became widespread, as shown by
the unequal sex ratio for the 1972--1984 period for women with two girls.

@@

% What has happened to spacing and why
The most substantial lengthening of birth intervals came from the best-educated women,
likely because of substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio reached almost 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the lengthening in the long intervals came immediately after the introduction of sex 
selection in India.

Some of the lengthening in birth intervals are so large that we even observed a reversal of
the traditional spacing pattern for some groups: 
For these groups, when there are no sons, we now see statistically significant longer,
rather than statistically significant shorter, birth intervals than if there is at least
one son.
The women who are the least likely to use sex selection, as indicated by their sex ratios
at birth, still show the traditional spacing pattern with short spacing in the absence of
sons.

@@

There are two critical questions that future research should address.
First, access to sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection is available could be an 
unintended side-effect, rather than the result of girls becoming more valued as is often 
assumed \citep{Hu2015}. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex selection.

@@

I use two approaches to examine the degree of recall error.
The first approach is to test whether the observed sex ratio is significantly different
from the natural sex ratio.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
Prenatal sex determination techniques did not become widely available until the mid-1980s, 
so any significant deviation from the natural sex ratio before that time is likely the 
result of recall error.
The second approach is to compare births that took place during the same period but
where captured in different surveys.
Recall error is likely to increase with time, so births and deaths that took place earlier 
are more likely to be subject to recall error than more recent events.

@@

Each predicted percent of boys is tested against the natural percentage of
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
The predicted percentage boys may differ from the natural rate because of 
natural variation, any remaining recall error not corrected for, or 
sex selection.

>>>>

\noindent Following the arrival of prenatal sex-determination technologies in the mid-1980s, 
India has experienced an increasingly male-biased sex ratio at birth, presumably from sex-selective 
abortions. 
Abortions lengthen birth intervals, but we know little about how birth spacing has changed or the 
effects of these changes. 
This paper shows that, although the length of birth intervals increased overall, well-educated women 
with no sons had the most substantial lengthening---and the most male-biased sex ratios. 
Furthermore, most of these changes took place immediately after the introduction of prenatal 
sex-determination technologies. 
Consequently, some women without sons now have longer birth intervals than those with sons, reversing 
India's traditional spacing pattern. 
Less-educated women continue short birth spacing when they have no sons, with only limited evidence 
of male-biased sex ratios. 
Because of the rapid lengthening of birth intervals, period fertility rates substantially overestimated 
how fast cohort fertility fell. 
Moreover, predicted cohort fertility is still 10\%--20\% above the period fertility rate. 
If lengthening of birth intervals arise from repeated abortions, the associated short pregnancy spacing 
may counteract any positive effects of longer birth spacing. 
There is, however, no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.

@@

An increased desire to invest in children may also lengthen birth spacing. 
The clearest example of a positive effect of longer spacing is for health as very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

@@

Even though higher potential wages from more education should shorten birth spacing and
lower fertility, the low and declining female labor force participation suggests little
economic incentive to space children closer together.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

@@

However, higher female education might still be associated with lower fertility and more 
sex selection, provided better-educated women are better at producing child human capital, 
as suggested by \citet{Behrman1999}. 
In a situation with increasing returns to male education, households may desire 
better-educated women because of their ability to produce better-educated sons, 
rather than because of their higher potential income. 
The higher return to male education provides an additional incentive to invest in sons' 
education---with a better-educated mother one such investment---and leads to lower desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.
The implication is that the usual policy recommendation of increased female education may not
change sex selection unless there is a concurrent increase in the relative return to female
education in the labor market.

@@

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated as their
household income increase the most, and because of their presumptive use of sex selection. 
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

@@

Second, even without sex selection, it is unlikely that characteristics, such as previous
births' sex composition, have the same effects throughout the entire birth interval.
The proportional hazard model requires that an individual's hazard is a 
fixed proportion of the hazard for any other individual. 
Nonconstant effects violate that assumption, and the results from a proportional hazard 
model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals because there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008}.

@@

Therefore, I use a nonproportional hazard specification that allows the shape of the 
hazard functions to vary across groups.
Furthermore, the combination of a nonproportional specification and a flexible baseline
hazard mitigates the potential effects of unobserved heterogeneity \citep{Dolton1995}.

@@

I divide the explanatory variables into two groups, nonproportional and proportional. 
The first group consists of characteristics shown in the prior literature to affect the 
spacing choice and the use of sex selection: Mother’s education, sex composition of 
previous children, and area of residence.
To minimize any potential bias from including proportional variables, I estimate separate 
models for each birth interval, education group, and period combination, rather than 
including education as a variable.

@@

Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentile birth interval durations in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with 12 or more years of 
education, and the fourth spell for women with 12 or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2\% boys
\citep{jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors and tests for statistically
significant differences across sex compositions are available in the online appendix.

@@

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth interval durations, 
where the more the parity progression probabilities declined, the more the birth interval 
lengthened. 
For example, the probability of a fourth birth for urban women with 8--11 years 
of education and two sons and one girl has declined by almost 40 percentage points as the 
75th percentile birth interval length increased by 22 months.
Compare this with rural women of no education with a boy as their first child, for whom the 
probability of a third birth declined by fewer than six percentage points while the birth 
interval increased by only slightly more than two months.

@@

Even more striking is that most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These birth spacing changes may even understate the impact because this particular group
appears to have had access to sex selection even before it became widespread, as shown by
the unequal sex ratio for the 1972--1984 period for women with two girls.

@@

% What has happened to spacing and why
The most substantial lengthening of birth intervals occurred among the best-educated women,
likely because of substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio reached almost 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the lengthening in the long intervals came immediately after the introduction of sex 
selection in India.

For some groups, the lengthening in birth intervals was so substantial that it led to a
reversal of the traditional spacing pattern: In the absence of sons, these groups now show
statistically significant longer, rather than statistically significant shorter, birth
intervals than if there is at least one son.

The women who are the least likely to use sex selection, as indicated by their sex ratios
at birth, still show the traditional spacing pattern with short spacing in the absence of
sons.

@@

There are two crucial questions that future research should address.

First, what is behind the improvements in girls' health status?
Access to sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection is available could be an 
unintended side-effect, rather than the result of girls becoming more valued as is often 
assumed \citep{Hu2015}. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex 
selection.

@@

I use two approaches to examine the degree of recall error.
The first approach is to test whether the observed sex ratio is significantly different
from the natural sex ratio.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{jacobsen99,Portner2015b}.
Prenatal sex determination techniques did not become widely available until the mid-1980s, 
so any significant deviation from the natural sex ratio before that time is likely the 
result of recall error.
The second approach is to compare births that took place during the same period but
where captured in different surveys.
Recall error is likely to increase with time, so births and deaths that took place earlier 
are more likely to be subject to recall error than more recent events.

@@

Each predicted percent of boys is tested against the natural percentage of
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{jacobsen99,Portner2015b}.
The predicted percentage boys may differ from the natural rate because of 
natural variation, any remaining recall error not corrected for, or 
sex selection.
